---
title: "Exploring and recalculating AFs<br>additional samples and variants filtering<br>adding AFs for subgtoups: wecare feb2016"
author: "Alexey Larionov"
output: html_document
---

started: Mar2016
last updated: 26Jun2016

# start_section

```{r start_section}

# Start time
Sys.time()

# Clean-up
rm(list=ls())
graphics.off()

# Set root working folder
library(knitr)
opts_knit$set(root.dir = "/scratch/medgen/scripts/rscripts_05.16")
#setwd("/scratch/medgen/scripts/rscripts_05.16")

library(dplyr) # for piping, filter, select etc

```

# load_and_check_data

```{r load_and_check_data}

load(file="data/s05_reshape_data_feb2016.RData")

ls()

dim(genotypes.mx)
class(genotypes.mx)
genotypes.mx[1:25,1:5]

dim(phenotypes.df)
colnames(phenotypes.df)
phenotypes.df[1:5,1:5]

dim(variants.df)
colnames(variants.df)
variants.df[1:5,1:5]

# check consistency of samples in genotypes and phenotypes
genotype_samples <- colnames(genotypes.mx)
length(genotype_samples)

all_phenotype_samples <- 
  phenotypes.df %>% select(wes_id)
all_phenotype_samples <- as.vector(all_phenotype_samples[,1])
length(all_phenotype_samples)

pass_phenotype_samples<- 
  phenotypes.df %>% filter(filter=="pass") %>% select(wes_id)
pass_phenotype_samples <- as.vector(pass_phenotype_samples[,1])
length(pass_phenotype_samples)

sum( genotype_samples %in% all_phenotype_samples )
sum( genotype_samples %in% pass_phenotype_samples )

rm(genotype_samples, all_phenotype_samples)

# check consistency of variants in genotypes and variants
sum(rownames(genotypes.mx) != variants.df$VarID )

```

# add_rownames_and_remove_failed_cases

```{r add_rownames_and_remove_failed_cases}

genotypes.mx <- genotypes.mx[,pass_phenotype_samples]
dim(genotypes.mx)

rownames(phenotypes.df) <- phenotypes.df$wes_id
phenotypes.df <- phenotypes.df[pass_phenotype_samples,]
dim(phenotypes.df)

rownames(variants.df) <- variants.df$VarID

rm(pass_phenotype_samples)

```

# explore_AF_AN_and_AC_in_whole_dataset

```{r explore_AF_AN_and_AC_in_whole_dataset}

# Explore AF, AC and AN values reported in vcf file
hist(variants.df$AF, breaks=seq(0, 1, 0.01))
hist(variants.df$AC, breaks=seq(0, 1030, 10))
hist(variants.df$AN, ylim=c(0,20000), labels=TRUE)

```

# recalculate_AF_AN_and_AC_in_whole_dataset

```{r recalculate_AF_AN_and_AC_in_whole_dataset}

# Function to calculate AN from genotypes
AN.fn <- function(x) 2*sum(!is.na(x))

# Calculate AF, AC and AN from genotypes
AC_GT <- apply(genotypes.mx, 1, sum, na.rm=TRUE)
AN_GT <- apply(genotypes.mx, 1, AN.fn)
AF_GT <- AC_GT / AN_GT

# Compare AF, AC and AN calculated from genotypes 
# with AF, AC and AN  values reported in vcf

plot(variants.df$AC, AC_GT, main="AC: vcf vs. genotypes")
abline(0,1,col="red")
# In most cases AC from genotypes is below than from vcf
# This is consistent with removal of some genotypes and whole cases during filtering

plot(variants.df$AN, AN_GT, 
     xlim=c(0,1000), ylim=c(0,1000), 
     main="AN: vcf vs. genotypes (range 0-1000)")
abline(0,1,col="red") 
# AN from genotypes is always smaller than from vcf
# No cases with llow AN because of the call-rate filter 
# and other filters applied earlier

plot(variants.df$AN, AN_GT, 
     main="AN: vcf vs. genotypes")
abline(0,1,col="red") 

plot(variants.df$AF, AF_GT, main="AF: vcf vs. genotypes") 
abline(0,1,col="red") 
# The AFs are concordant, except for some variants, 
# which will be removed below

# Add gt AF_GT, AN_GT, AC_GT to variants.df
variants.df <- cbind(variants.df, AF_GT, AC_GT, AN_GT)
hist(variants.df$AF_GT, breaks=seq(0, 1, 0.01))
hist(variants.df$AC_GT, breaks=seq(0, 1030, 10))
hist(variants.df$AN_GT, ylim=c(0,5500), labels=TRUE)

# Clean-up
rm(AF_GT, AC_GT, AN_GT)

```

# exclude_variants_with_zero_AF_after_filtering

```{r exclude_variants_with_zero_AF_after_filtering}

variants_to_exclude <- variants.df$AF_GT == 0
sum(variants_to_exclude)

genotypes.mx <- genotypes.mx[ !variants_to_exclude, ]
dim(genotypes.mx)

variants.df <- variants.df[ !variants_to_exclude, ]
dim(variants.df)

rm(variants_to_exclude)

```

# exclude_variants_with_discordant_AF_vs_AC_AN
Is this needed ?

```{r exclude_variants_with_discordant_AF_vs_AF-GT}

# Variants with discrepancies between AF estimates in VCF file
variants_to_exclude <- 
  variants.df$AF_GT - variants.df$AF > 0.01 | 
  variants.df$AF_GT - variants.df$AF < -0.01
sum(variants_to_exclude)

#table(as.vector(variants.df[variants_to_exclude, "SYMBOL"]))

# The highest 3 discrepancies
as.vector(variants.df[variants.df$AF_GT - variants.df$AF > 0.05, "SYMBOL"])
as.vector(variants.df[variants.df$AF_GT - variants.df$AF < -0.05, "SYMBOL"])

# Exclude variants with discrepancies
genotypes.mx <- genotypes.mx[ !variants_to_exclude, ]
dim(genotypes.mx)

variants.df <- variants.df[ !variants_to_exclude, ]
dim(variants.df)

rm(variants_to_exclude)

```

# explore_AF-GT_AC-GT_AN-GT_after_additional_filtering

```{r explore_AF-GT_AC-GT_AN-GT_after_additional_filtering}

hist(variants.df$AF_GT, breaks=seq(0, 1, 0.01))
hist(variants.df$AC_GT, breaks=seq(0, 1030, 10))
hist(variants.df$AN_GT, ylim=c(0,5500), labels=TRUE)

plot(variants.df$AF, variants.df$AF_GT, main="Corrected AF: vcf vs. genotypes") 
abline(0,1,col="red") 
# The AFs concordance has improved, especially in rare variants

```

# Calculate AFs for Contralateral BC (CBC, cases)

```{r calculate_cbc_afs}

# Select cbc cases
# 1=Case, 0=Control (Confirmed by DC in e-mail to EF of 29-30Mar2016)
cbc.idx <- phenotypes.df$cc == 1 
sum(cbc.idx)

gt.cbc.df <- genotypes.mx[,cbc.idx]
dim(gt.cbc.df)

# Calculate AF, AC and AN from genotypes
AC_CBC <- apply(gt.cbc.df, 1, sum, na.rm=TRUE)
AN_CBC <- apply(gt.cbc.df, 1, AN.fn)
AF_CBC <- AC_CBC / AN_CBC
hist(AF_CBC, breaks=seq(0,1,0.01))

variants.df <- cbind(variants.df, AF_CBC, AC_CBC, AN_CBC)
rm(gt.cbc.df, cbc.idx, AF_CBC, AC_CBC, AN_CBC)

```

# Calculate AFs for Unilateral BC (UBC, controls)

```{r calculate_ubc_afs}

# Select cbc cases
# 1=Case, 0=Control (Confirmed by DC in e-mail to EF of 29-30Mar2016)
ubc.idx <- phenotypes.df$cc == 0 
sum(ubc.idx)

gt.ubc.df <- genotypes.mx[,ubc.idx]
dim(gt.ubc.df)

# Calculate AF, AC and AN from genotypes
AC_UBC <- apply(gt.ubc.df, 1, sum, na.rm=TRUE)
AN_UBC <- apply(gt.ubc.df, 1, AN.fn)
AF_UBC <- AC_UBC / AN_UBC
hist(AF_UBC, breaks=seq(0,1,0.01))

variants.df <- cbind(variants.df, AF_UBC, AC_UBC, AN_UBC)
rm(gt.ubc.df, ubc.idx, AF_UBC, AC_UBC, AN_UBC)

```

# Calculate AFs for Unilateral Sporadic UBC (USBC)

```{r calculate_usbc_afs}

# Select usbc cases
# coding: 1=Case/Familial, 0=Control/Sporadic 
#(Confirmed by DC in e-mail to EF of 29-30Mar2016)
ubc.idx <- phenotypes.df$cc == 0 
sum(ubc.idx)

sum(is.na(phenotypes.df$family_history))
not.na.fam_hist <- ! is.na(phenotypes.df$family_history)

sbc.idx <- phenotypes.df$family_history == 0 & not.na.fam_hist
sum(sbc.idx)

usbc.idx <- ubc.idx & sbc.idx
sum(usbc.idx)

#View(phenotypes.df[usbc.idx, c("cc","family_history")])
sum(phenotypes.df[usbc.idx, c("cc","family_history")])

gt.usbc.df <- genotypes.mx[,usbc.idx]
dim(gt.usbc.df)

# Calculate AF, AC and AN from genotypes
AC_USBC <- apply(gt.usbc.df, 1, sum, na.rm=TRUE)
AN_USBC <- apply(gt.usbc.df, 1, AN.fn)
AF_USBC <- AC_USBC / AN_USBC
hist(AF_USBC, breaks=seq(0,1,0.01))

variants.df <- cbind(variants.df, AF_USBC, AC_USBC, AN_USBC)
rm(not.na.fam_hist, gt.usbc.df, ubc.idx, sbc.idx, usbc.idx, AF_USBC, AC_USBC, AN_USBC)

```

# Calculate AFs for Bilateral Familial BC (CFBC)

```{r calculate_cfbc_afs}

# Select cfbc cases
# coding: 1=Case/Familial, 0=Control/Sporadic 
#(Confirmed by DC in e-mail to EF of 29-30Mar2016)
cbc.idx <- phenotypes.df$cc == 1 
sum(cbc.idx)

sum(is.na(phenotypes.df$family_history))
not.na.fam_hist <- ! is.na(phenotypes.df$family_history)

fbc.idx <- phenotypes.df$family_history == 1 & not.na.fam_hist
sum(fbc.idx)

cfbc.idx <- cbc.idx & fbc.idx
sum(cfbc.idx)

#View(phenotypes.df[cfbc.idx,c("cc","family_history")])
sum(phenotypes.df[cfbc.idx,c("cc","family_history")])

gt.cfbc.df <- genotypes.mx[,cfbc.idx[!is.na(cfbc.idx)]]
dim(gt.cfbc.df)

# Calculate AF, AC and AN from genotypes
AC_CFBC <- apply(gt.cfbc.df, 1, sum, na.rm=TRUE)
AN_CFBC <- apply(gt.cfbc.df, 1, AN.fn)
AF_CFBC <- AC_CFBC / AN_CFBC
hist(AF_CFBC, breaks=seq(0,1,0.01))

variants.df <- cbind(variants.df, AF_CFBC, AC_CFBC, AN_CFBC)
rm(not.na.fam_hist, gt.cfbc.df, cbc.idx, fbc.idx, cfbc.idx, AF_CFBC, AC_CFBC, AN_CFBC, AN.fn)

```

# save_data

```{r save_data}

save.image(file="data/s06_recalculate_afs_feb2016.RData")

```

# final_section

```{r final_section}

sessionInfo()
Sys.time()

```
